Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-09-24 Paper/abstract.tex
update ...

\section*{Abstract}
We focus on efficient quantification of uncertainty in complex chemical
reaction networks with a large number of uncertain parameters.  Parameter
dimension reduction is accomplished by computing an~\emph{active subspace} that
predominantly captures the variability in the quantity of interest (QoI).  In
this study, we compute the active subspace for the H$_2$/O$_2$ mechanism that
involves 19 chemical reactions, using an efficient iterative strategy.  The
active subspace is first computed for a 19-parameter problem wherein only the
uncertainty in the pre-exponents of the individual reaction rates is
considered. This is followed by the analysis of a 33-dimensional case wherein
the activation energies are also considered uncertain.  In both cases, a
one-dimensional active subspace is identified, which indicates
enormous potential for efficient statistical analysis of complex 
chemical systems. In addition, we explore links between active subspaces and
global sensitivity analysis measures, and exploit these links for 
indentification of key contributors to the variability in the model response.

%In this case, the two approaches are observed to yield consistent results.
%However, when computing the active subspace for a 33-dimensional problem
%involving uncertain activation energies in addition to the pre-exponents, a
%trade-off between accuracy and effort is observed.

%Specifically, when estimating the gradient using
%the regression-based approach, the variability of the QoI (in the subspace),
%due to uncertainty in the rate parameters, is under-estimated. Moreover, the
%sensitivity towards one of the most important parameter is not captured. Hence,
%we carefully recommend using the regression-based approach for estimating the
%gradient only in situations involving intensive computations of the model 
%(e.g. complex mechanisms involving a large number of reactions and species), and where the
%goal is to obtain rough estimates of the statistics of the QoI.

>>>>

\section*{Abstract}
We focus on efficient quantification of uncertainty in complex chemical
reaction networks with a large number of uncertain parameters.  Parameter
dimension reduction is accomplished by computing an~\emph{active subspace} that
predominantly captures the variability in the quantity of interest (QoI).  In
the present work, we compute the active subspace for a H$_2$/O$_2$ mechanism 
that involves 19 chemical reactions, using an efficient iterative strategy.  The
active subspace is first computed for a 19-parameter problem wherein only the
uncertainty in the pre-exponents of the individual reaction rates is
considered. This is followed by the analysis of a 33-dimensional case wherein
the activation energies are also considered uncertain.  In both cases, a
one-dimensional active subspace is identified, which indicates enormous
potential for efficient statistical analysis of complex chemical systems. In
addition, we explore links between active subspaces and global sensitivity
analysis, and exploit these links for identification of key
contributors to the variability in the model response.

### 2018-09-24 Paper/conc.tex
update ...

%Additionally, the surrogates were assessed by comparing probability
%distributions for the ignition delay,
%constructed using surrogate predictions and model evaluations at the same set
%of 10$^4$ random samples. The PDFs based on model evaluations and the
%1-dimensional surrogate from the perturbation-based approach were found to be
%almost identical. However, the PDF based on 1-dimensional surrogate from the
%regression-based approach, although captured the modal estimate with reasonable
%accuracy, was observed to underestimate the uncertainty in the ignition delay.
%Mean estimates of the ignition delay using model evaluations and predictions
%using the two surrogates at 10$^4$ samples were found to be in agreement.
%However, the standard deviation, computed using predictions from
%the surrogate resulting from the regression-based approach was found to be accurate
%only upto the first significant digit. Moreover, it was observed that the
%activity scores, computed using the regression-based strategy failed to capture
%the sensitivity towards one of the most important parameters, $E_{a,15}$.  
%
%More specifically, using
%the regression-based approach, 
%\begin{itemize} 
%\item The desired statistics (e.g. mean and variance) of the
%QoI might not be approximated to a desired level of accuracy.  
%\item The
%approach might fail to capture the sensitivity  of the QoI towards all inputs.
%\end{itemize} 
Therefore, in general, perturbation-based approaches should be preferred. The
regression-based approach can be explored in situations involving intensive
simulations where the perturbation-based methods are intractable, and the goal
is obtaining rough estimates of the statistics of the QoI as opposed to a more
detailed analysis such as those involving parametric sensitivity.  We also
mention that alternate regression based approaches such as ones based on
computing a global quadratic model have been proposed and used in the
literature; see e.g.,~\cite{Constantine:2017a}. The applicability of such an
approach in the context of high-dimensional chemical reaction networks is
subject to future work.

The computational framework presented in this work is agnostic to the choice of
the chemical system and can be easily adapted for other systems as long
as the quantity of interest is continuously differentiable in the considered
domain of the inputs.  We have demonstrated that the active subspace could be
exploited for efficient forward propagation of the uncertainty from inputs to
the output. However, the activity scores and the low-dimensional surrogate
could further guide optimal allocation of computational resources for
calibration of the important rate-controlling parameters in a Bayesian setting.
Additionally, dimension reduction using active subspaces could assist in
developing robust formulations for predicting discrepancy between simulations
and measurements due to epistemic uncertainty in the model inputs.

>>>>

%Additionally, the surrogates were assessed by comparing probability
%distributions for the ignition delay,
%constructed using surrogate predictions and model evaluations at the same set
%of 10$^4$ random samples. The PDFs based on model evaluations and the
%1-dimensional surrogate from the perturbation-based approach were found to be
%almost identical. However, the PDF based on 1-dimensional surrogate from the
%regression-based approach, although captured the modal estimate with reasonable
%accuracy, was observed to underestimate the uncertainty in the ignition delay.
%Mean estimates of the ignition delay using model evaluations and predictions
%using the two surrogates at 10$^4$ samples were found to be in agreement.
%However, the standard deviation, computed using predictions from
%the surrogate resulting from the regression-based approach was found to be accurate
%only upto the first significant digit. Moreover, it was observed that the
%activity scores, computed using the regression-based strategy failed to capture
%the sensitivity towards one of the most important parameters, $E_{a,15}$.  
%
%More specifically, using
%the regression-based approach, 
%\begin{itemize} 
%\item The desired statistics (e.g. mean and variance) of the
%QoI might not be approximated to a desired level of accuracy.  
%\item The
%approach might fail to capture the sensitivity  of the QoI towards all inputs.
%\end{itemize} 
Therefore, in general, perturbation-based approaches should be preferred for 
active subspace computation.
The regression-based approach can be explored in situations involving intensive
simulations where the perturbation-based methods are intractable.
%
%and the goal is obtaining rough estimates of the statistics of the QoI as
%opposed to a more detailed analysis such as those involving parametric
%sensitivity.  
%
We also mention that alternate regression based approaches such as ones based
on computing a global quadratic model have been proposed and used in the
literature; see e.g.,~\cite{Constantine:2017a}.  The applicability of such an
approach in the context of high-dimensional chemical reaction networks is
subject to future work.

The computational framework presented in this work is agnostic to the choice of
the chemical system and can be easily adapted for other systems as long
as the quantity of interest is continuously differentiable in the considered
domain of the inputs.  We have demonstrated that the active subspace could be
exploited for efficient forward propagation of the uncertainty from inputs to
the output. Moreover, the activity scores and the low-dimensional surrogate
could further guide optimal allocation of computational resources for
calibration of the important rate-controlling parameters in a Bayesian setting.
Additionally, dimension reduction using active subspaces could assist in
developing robust formulations for predicting discrepancy between simulations
and measurements due to epistemic uncertainty in the model inputs.
